Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-08-15 manuscript.tex
ms edits

\subsection*{Y-chromosome diversity in \textit{R. hastatulus} is very low}
Our analysis revealed that diversity on the \textit{R. hastatulus} Y chromosome is significantly lower than expected under neutrality, with estimates indicating $Y/A=0.02$, which is 12.5 fold lower than the standard neutral prediction of $Y/A = 0.25 (P<0.0001)$, and a 40-fold reduction compared to mean diversity on the X chromosome (Table 1). Note that by normalizing X and Y diversity by autosomal diversity, our results indicate that the X-Y difference we observed was not due to an elevation of X chromosome diversity, but rather a Y-specific reduction. Conceivably, such low diversity on the Y could arise at least in part from a low mutation rate on the Y, or a lower mutation rate in males compared to females. However, these possibilities can be rejected because there is no evidence that the number of synonymous mutations in X and Y lineages, estimated by both parsimony and maximum likelihood, are significantly different \citep{hough2014}. In addition, the pattern of reduced Y chromosome diversity occurs in each of the \textit{R. hastatulus} sub-clades that were inferred to be monophyletic (\X Figure 2; Table 1; Figure S1,S2), suggesting that the effect is not population-specific.

@@

%In contrast to the Texas sex chromosome race, our analysis revealed that Y-chromosome diversity was considerably higher in the North Carolina race (see Supplementary Material Figure S1). This difference could have arisen due to population substructure, resulting in an elevated overall Y diversity in the pooled sample. To test this hypothesis, we conducted a phylogenetic analysis of sex-linked sequences from all populations in our study. The resulting sequence phylogeny (Figure 4) revealed strong support for the Y-linked genes from the XY1Y2 race being paraphyletic (98% bootstrap support). In particular, our analysis revealed that samples from two populations, both from South Carolina (hereafter the SC sub-clade), formed a monophyletic group, whereas samples from Florida and Georgia (hereafter the FL sub-clade) were also monophyletic, but were more closely related to the XY Texas race. Although our sampling here is limited, these results suggest that introgression may have occurred between the ancestral Texas (XY race) and the derived North Carolina (XY1Y2) race, leading to a second, derived, XY1Y2 sub-clade and an elevated level of Y-diversity in our pooled sample. As these races are inter-fertile (Smith reference about hybrid zone), we suggest that the SC sub-clade likely originated through hybridization between a female from the FL clade harboring the X-autosome fusion, and a male from the XY race. Notably, when we separated samples by the two inferred sub-clades of Y–linked sequences, the estimated level of Y diversity was consistent across all three clades, suggesting that the extensive reduction in diversity on this chromosome is not a population or race-specific phenomenon (see Supplementary Material, Figure S1).

>>>>

\subsection*{Y-chromosome diversity in \textit{R. hastatulus} is very low}
Our analysis revealed that diversity on the \textit{R. hastatulus} Y chromosome is significantly lower than expected under neutrality, with estimates indicating $Y/A=0.02$, which is 12.5 fold lower than the standard neutral prediction of $Y/A = 0.25 (P<0.0001)$, and a 40-fold reduction compared to mean diversity on the X chromosome (Table 1). Note that by normalizing X and Y diversity by autosomal diversity, our results indicate that the X-Y difference we observed was not due to an elevation of X chromosome diversity, but rather a Y-specific reduction. Conceivably, such low diversity on the Y could arise from a low mutation rate on the Y chromosome, or a lower mutation rate in males compared to females. However, these possibilities can be rejected because there is no evidence that the number of synonymous mutations in X and Y lineages, estimated by both parsimony and maximum likelihood, are significantly different \citep{hough2014}. It is worth noting that our results only apply to the $Y_{1}$ chromosome, however, as estimates of diversity were calculated using the set of sex-linked genes that were shared between the $XX/XY$ and the $XX/XY_{1}Y_{2}$ sex chromosome races. Previous work investigating patterns of genetic degeneration in both systems found that both the $Y_{1}$ and the more recently evolved "neo" $Y_{2}$ chromosomes have undergone classic signs of degeneration, including gene loss, loss of expression, and an accumulation of amino acid-changing mutations, but we cannot say at present whether the neo-Y has also undergone a significant reduction in diversity. Nonetheless, our results indicate that the pattern of reduced diversity on the $Y_{1}$ chromosome occurs in each of the \textit{R. hastatulus} sub-clades that were inferred to be monophyletic (\X Figure 2; Table 1; Figure S1,S2), suggesting that this effect is not population-specific.

@@

Although our sampling of each \textit{R. hastatulus} sub-clade is limited, the discovery of three phylogenetically distinct monophyletic groups is interesting as it suggests the possibility that introgression occurred between the ancestral Texas (XY race) and the derived North Carolina ($XY_{1}Y_{2}$) race, leading to a derived $XY_{1}Y_{2}$ sub-clade. As the North Carolina and Texas races are known to be inter-fertile \citep{smith1964evolving}, we suggest that the SC sub-clade inferred here likely originated through hybridization between a female from the FL clade harboring the X-autosome fusion, and a male from the XY Texas race. Notably, when we pooled the samples from each sub-clade together, the estimated level of Y diversity was significantly higher in the pooled data, highlighting the presence of strong substructure between these groups (Figure S2).
